---
title:  Trains a one-class dense model on the PCBC dataset
author: Artem Sokolov
output: html_document 
---
***
#Purpose 

***
# Introduction


# Setup 
Make sure you have the following packages downloaded and loaded into your library. 
We will retrieve data from Synapse, an open source data platform. [synapse](https://www.synapse.org/)
A free account must be made before accessing synapse data, running the 'synapseLogin' function will prompt you to loin.
```{r,error= FALSE}
library(gelnet)
library(dplyr)
library(biomaRt)
library(synapseClient)
synapseLogin()

```



# genes2hugo function
Ensembl is a gene browser that will be used to gather information on the gene type and geneid. [ensembl](www.ensembl.org)
useMArt is the function used connect to the Ensembl website and select the dataset you with to work with. 
Use the getBM function to retrieve the the data for the SRC gene from ensembl.
```{r, eval = FALSE}
  ## Retrieve the EMSEMBL -> HUGO mapping
  ensembl <- biomaRt::useMart( "ENSEMBL_MART_ENSEMBL", host="www.ensembl.org", dataset="hsapiens_gene_ensembl" )
  ID <- biomaRt::getBM( attributes=c('srcType', 'hgnc_symbol'), filters=srcType, values=v, mart=ensembl )

```

Next make sure the data was mapped by checking that there is at least one row of data.
Filter out empty data
```{r, eval = FALSE}
 ID <- biomaRt::getBM( attributes=c(srcType, "hgnc_symbol"), filters=srcType, values=v, mart=ensembl )
   ## Make sure there was at least one mapping
  if( nrow(ID) < 1 )
    stop( "No IDs mapped successfully" )
  
  ## Drop empty duds
  j <- which( ID[,2] == "" )
  if( length(j) > 0 )
    ID <- ID[-j,]
  stopifnot( all( ID[,1] %in% v ) )
  
  ID
```

##Continuous genes2hugo Function
```{r, error = FALSE, }
## Maps ENSEMBL IDs to HUGO
## Use srcType = "ensembl_gene_id" for Ensembl IDs
## Use srcType = "entrezgene" for Entrez IDs
genes2hugo <- function( v, srcType = "ensembl_gene_id" )
{
  ## Retrieve the EMSEMBL -> HUGO mapping
  ensembl <- biomaRt::useMart( "ENSEMBL_MART_ENSEMBL", host="www.ensembl.org", dataset="hsapiens_gene_ensembl" )
  ID <- biomaRt::getBM( attributes=c(srcType, "hgnc_symbol"), filters=srcType, values=v, mart=ensembl )
   ## Make sure there was at least one mapping
  if( nrow(ID) < 1 )
    stop( "No IDs mapped successfully" )
  
  ## Drop empty duds
  j <- which( ID[,2] == "" )
  if( length(j) > 0 )
    ID <- ID[-j,]
  stopifnot( all( ID[,1] %in% v ) )
  
  ID
}

```


#Train function 

Here is where synapse will be used to retrieve the information. 
[synGet](https://www.rdocumentation.org/packages/synapseClient/versions/1.13-4/topics/synGet)  will be used to load and save RNAseq data.
[synTableQuery](https://www.rdocumentation.org/packages/synapseClient/versions/1.13-4/topics/synTableQuery) is used to gather meta data. 
In this case the downlaod location is set to '/data/PCBC' but it can be changed to wherever you would like or it can be set to 'NULL' and it will be stored in your chache.

'X' is a large matrix 13189 genes by 229 cell samples.
'Y' is a single variable data frame consting of 301 observations. 
```{r}
  ## Load RNAseq data
  synRNA <- synGet( "syn2701943", downloadLocation = "/data/PCBC" )
  X <- read.delim( synRNA@filePath ) %>%
    tibble::column_to_rownames( "tracking_id" ) %>%
    as.matrix()
  X[1:3,1:3]

```

```{r}
  ## Retrieve metadata
  synMeta <- synTableQuery( "SELECT UID, Diffname_short FROM syn3156503" )
  Y <- synMeta@values %>%
    mutate( UID = gsub("-", ".", UID) ) %>%
    tibble::column_to_rownames( "UID" )
  head(Y)
```

 Filter the labels from the meta data. Notice the change in format for 'y' 
```{r}
  ## Retrieve the labels from the metadata
  y <- Y[colnames(X),]
 
  names(y) <- colnames(X)
  ## Fix the missing labels by hand
  y["SC11.014BEB.133.5.6.11"] <- "EB"
  y["SC12.039ECTO.420.436.92.16"] <- "ECTO"
  
  ## Drop the splice form ID from the gene names
  v <- strsplit( rownames(X), "\\." ) %>% lapply( "[[", 1 ) %>% unlist()
  
  rownames(X) <- v
  head(y)
  
```

Call the genes2hugo function created above and use the labels from the RNSseq data to map genetype and geneid.
The result of this function is a data frame with two variables consisting of 12952 observations.
```{r}
 ## Map Ensembl IDs to HUGO
  V <- genes2hugo( rownames(X) )
  head(V)
 
```


Now change the row names of X from the gene name to the hgnc (HUGO Gene Nomenclature Committee) symbol. Notice that the dimensions of 'X' change from 13189 rows to 12952 rows.
```{r}
 X <- X[V[,1],]
  rownames(X) <- V[,2]
  X[1:3,1:3]
```
```{r}
  dim(X)
```

 Reduce gene set to the provides list.
```{r,eval= FALSE}
  if( is.null( fnGenes ) == FALSE )
  {
    vGenes <- read.delim( fnGenes, header=FALSE ) %>% as.matrix() %>% drop()
    VE <- genes2hugo( vGenes, "entrezgene" )
    X <- X[intersect( rownames(X), VE[,2] ),]
  }
```

Find the mean center by subtracting the mean of each gene (m) from the RNSseq data(X).
```{r}
  m <- apply( X, 1, mean )
  X <- X - m
  X[1:3,1:3]

```
Identify stem cells and break up all samples into 2 groups:
*Stem cell(X.tr) -- 78 samples 
*not stem cell (X.bk) -- 151 samples 
```{r}
  j <- which( y == "SC" )
  X.tr <- X[,j]
  X.tr[1:3,1:3]
```
```{r}
  X.bk <- X[,-j]
  X.bk[1:3,1:3]
```


Now we can begin to train the the one-class model with the gelnet function using the PCBC stem cell data. 
The [gelnet] (https://www.rdocumentation.org/packages/gelnet/versions/1.2.1/topics/gelnet) function can be used for Linear Regression, Binary Classification and One class Problems by using an iterative method called coordinanted descent.

gelnet(X, y, l1, l2)
X: n by p matrix => transpose( 'X.r' )
y: NULL for one class models 
l1: coefficient for the L1-norm penalty => '0' 
l2: coefficient for the L2-norm penalty => '1'

Make sure you transpose the matrix so that the genes are listed as rows and samples as columns.
Then store the results as a tsv file (pcbc-stemsig.tsv). 
```{r, error= FALSE }
  mm <- gelnet( t(X.tr), NULL, 0, 1 )
  
```
```{r, eval= FALSE}
  write.table( mm$w, file=fnOut, sep="\t", quote=FALSE, col.names=FALSE )
  
```


Leave-one-out cross validation will be used to test the accuracy of the model. Running this chunk will return a numeric vector containing the same size as the number of samples of the 'X.tr' variable (in this case 78). If it is successful the returned values should all be equal to 1.
```{r, eval = FALSE }
  ## Perform leave-one-out cross-validation
  auc <- c()
  for( i in 1:ncol(X.tr) )
  {
    ## Train a model on non-left-out data
    X1 <- X.tr[,-i]
    m1 <- gelnet( t(X1), NULL, 0, 1 )
    
    ## Score the left-out sample against the background
    s.bk <- apply( X.bk, 2, function(z) {cor( m1$w, z, method="sp" )} )
    s1 <- cor( m1$w, X.tr[,i], method="sp" )
    
    ## AUC = P( left-out sample is scored above the background )
    auc[i] <- sum( s1 > s.bk ) / length(s.bk)
    cat( "Current AUC: ", auc[i], "\n" )
    cat( "Average AUC: ", mean(auc), "\n" )
  }
  
  return(auc)
```


##The continous main.train Function
```{r, eval= FALSE}
## fnOut - filename of the output signature
## fnGenes - [optional] filename of the list of entrez ID to consider
main.train <- function( fnOut = "pcbc-stemsig.tsv", fnGenes = NULL )
{
  ## Load RNAseq data
  synRNA <- synGet( "syn2701943", downloadLocation = "/data/PCBC" )
  X <- read.delim( synRNA@filePath ) %>%
    tibble::column_to_rownames( "tracking_id" ) %>%
    as.matrix()
  
  ## Retrieve metadata
  synMeta <- synTableQuery( "SELECT UID, Diffname_short FROM syn3156503" )
  Y <- synMeta@values %>%
    mutate( UID = gsub("-", ".", UID) ) %>%
    tibble::column_to_rownames( "UID" )
  
  ## Retrieve the labels from the metadata
  y <- Y[colnames(X),]
  names(y) <- colnames(X)
  
  ## Fix the missing labels by hand
  y["SC11.014BEB.133.5.6.11"] <- "EB"
  y["SC12.039ECTO.420.436.92.16"] <- "ECTO"
  
  ## Drop the splice form ID from the gene names
  v <- strsplit( rownames(X), "\\." ) %>% lapply( "[[", 1 ) %>% unlist()
  rownames(X) <- v
  
  ## Map Ensembl IDs to HUGO
  V <- genes2hugo( rownames(X) )
  X <- X[V[,1],]
  rownames(X) <- V[,2]
  
  ## Reduce the gene set to the provided list (if applicable)
  if( is.null( fnGenes ) == FALSE )
  {
    vGenes <- read.delim( fnGenes, header=FALSE ) %>% as.matrix() %>% drop()
    VE <- genes2hugo( vGenes, "entrezgene" )
    X <- X[intersect( rownames(X), VE[,2] ),]
  }
  
  ## Mean-center the data
  m <- apply( X, 1, mean )
  X <- X - m
  
  ## Identify stem cell samples
  j <- which( y == "SC" )
  X.tr <- X[,j]
  X.bk <- X[,-j]
  
  ## Train a one-class model
  mm <- gelnet( t(X.tr), NULL, 0, 1 )
  
  ## Store the signature to a file
  write.table( mm$w, file=fnOut, sep="\t", quote=FALSE, col.names=FALSE )
  
  ## Perform leave-one-out cross-validation
  auc <- c()
  for( i in 1:ncol(X.tr) )
  {
    ## Train a model on non-left-out data
    X1 <- X.tr[,-i]
    m1 <- gelnet( t(X1), NULL, 0, 1 )
    
    ## Score the left-out sample against the background
    s.bk <- apply( X.bk, 2, function(z) {cor( m1$w, z, method="sp" )} )
    s1 <- cor( m1$w, X.tr[,i], method="sp" )
    
    ## AUC = P( left-out sample is scored above the background )
    auc[i] <- sum( s1 > s.bk ) / length(s.bk)
    cat( "Current AUC: ", auc[i], "\n" )
    cat( "Average AUC: ", mean(auc), "\n" )
  }
  
  return(auc)
}
```


# Predict function

Use [read.delium] (http://rfunction.com/archives/1441) to load the signiture from the svaed file(pcbc-stemsig.tsv) and store it as a variable w. 
w is a numeric vector the same size as the number of selected genes, in this case 12952.
```{r}
  fnSig = "pcbc-stemsig.tsv"
  w <- read.delim( fnSig, header=FALSE, row.names=1 ) %>% as.matrix() %>% drop()
  
```
```{r}
w[1:10]
```

Create a fucntion f that will be used in the next steps to ..
```{r}
  ## Reduces HUGO|POSITION gene IDs to just HUGO
  f <- function( v ) unlist( lapply( strsplit( v, "\\|" ), "[[", 1 ) )
  
```


Again we retrieve data from synapse. For this specific data you will need NIH approval to get access. 
Once you gather the data, create a data frame 'X' and filter so that 'X' contains genes that mapped and values from the signiture gene set. 
'X' will be a large data frame with 11810 rows by 11070 columns.
```{r}
  s <- synGet( "syn4976369", downloadLocation = "/data/pancan" )
  X <- read.delim( s@filePath, as.is=TRUE, check.names=FALSE ) %>%	## Read the raw values
    filter( !grepl( "\\?", gene_id ) ) %>%		## Drop genes with no mapping to HUGO
    mutate( gene_id = f( gene_id ) ) %>%		## Clip gene ids to HUGO
    filter( gene_id %in% names(w) )			## Reduce to the signature's gene set
  head(X)
  
```

The data for SLC35E2 will have multiple gene id's. Filter it by only keeping the first entry.
```{r}
  j <- grep( "SLC35E2", X[,1] )
  if( length(j) > 1 )
    X <- X[-j[-1],]
  
```

Convert the data frame 'X' to a matrix. Rows are Gene id's and columns are Gene Sets. Notice the the size of this matirx is now 11809 rows by 11069 columns. 
```{r}
  rownames(X) <- NULL
  X <- X %>% tibble::column_to_rownames( "gene_id" ) %>% as.matrix()
  X[1:3,1:3]
```

 Reduce the signature to the common set of genes.
```{r}
  stopifnot( all( rownames(X) %in% names(w) ) )
  w <- w[ rownames(X) ]
  w[1:5]
```


Score the Maxtix 'X' using Spearman coorelation. 
```{r}
  s <- apply( X, 2, function(z) {cor( z, w, method="sp", use="complete.obs" )} )
  s[1:5]
```

Scale the scores to be between 0 and 1
```{r}
  s <- s - min(s)
  s <- s / max(s)
  s[1:5]
  
```

Then output scores to a file mRNA_StemScore.tsv.
```{r, eval=FALSE}
  write.table( cbind(s), file=fnOut, sep="\t", quote=FALSE, col.names=FALSE )
```


##The continous main.predict Function 
```{r}
## Uses the signature stored in fnSig to score PanCan33 data and stores the result to fnOut
main.predict <- function( fnSig = "pcbc-stemsig.tsv", fnOut = "mRNA_StemScore.tsv" )
{
  ## Load the signature
  w <- read.delim( fnSig, header=FALSE, row.names=1 ) %>% as.matrix() %>% drop()
  
  ## Reduces HUGO|POSITION gene IDs to just HUGO
  f <- function( v ) unlist( lapply( strsplit( v, "\\|" ), "[[", 1 ) )
  
  s <- synGet( "syn4976369", downloadLocation = "/data/pancan" )
  X <- read.delim( s@filePath, as.is=TRUE, check.names=FALSE ) %>%	## Read the raw values
    filter( !grepl( "\\?", gene_id ) ) %>%		## Drop genes with no mapping to HUGO
    mutate( gene_id = f( gene_id ) ) %>%		## Clip gene ids to HUGO
    filter( gene_id %in% names(w) )			## Reduce to the signature's gene set
  
  ## SLC35E2 has multiple entries with the same HUGO id
  ## Keep the first entry only
  j <- grep( "SLC35E2", X[,1] )
  if( length(j) > 1 )
    X <- X[-j[-1],]
  
  ## Convert to a matrix
  rownames(X) <- NULL
  X <- X %>% tibble::column_to_rownames( "gene_id" ) %>% as.matrix()
  
  ## Reduce the signature to the common set of genes
  stopifnot( all( rownames(X) %in% names(w) ) )
  w <- w[ rownames(X) ]
  
  ####### Score via Spearman correlation
  s <- apply( X, 2, function(z) {cor( z, w, method="sp", use="complete.obs" )} )
  
  ## Scale the scores to be between 0 and 1
  s <- s - min(s)
  s <- s / max(s)
  
  write.table( cbind(s), file=fnOut, sep="\t", quote=FALSE, col.names=FALSE )
}

```



# Main Function 
Once you have created all of the 'genes2hugo' 'main.train' 'main.predict' adn 'f' create the 'main' function. This function will train and apply the full and reduced signatures. After you have have created all of the functions run 'main' to preform the full analysis. 
```{r}
main <- function()
{
  ## Train a full signature, which will be saved to pcbc-stemsig.tsv
  main.train()
  
  ## Apply the full signature to score the entire PanCan33 cohort
  main.predict()
  
  ## Train a reduced signature, using the provided list of entrez IDs
  main.train( "pcbc-stemsig-reduced.tsv", "neighbor_genes_list_1042.txt" )
  
  ## Apply the reduced signature to score the entire PanCan33 cohort
  main.predict( "pcbc-stemsig-reduced.tsv", "mRNA_StemScore_reduced.tsv" )
}
```


