---
title: "PanCanStem"
author: "Tathiane Maistro Malta (code/doc), Artem Sokolov (code/doc), Michael Wells (doc), Tiago Silva (doc)"
output:
  html_document:
    highlight: tango
    theme: journal
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: no
editor_options:
  chunk_output_type: inline
---

# Introduction

This document contains workflows, explaining how to download stem/progenitor cells from the Progenitor Cell Biology Consortium(PCBC) and TCGA PanCan33 datasets, train a stemness signature using normal stem cells, and apply it to score TCGA tumor samples. The mRNAsi section contains the RNA expression-based stemness index workflow; while the mDNAsi contains the DNA methylation-based stemness index workflow.


# Authors

The messenger RNA stemness index (mRNAsi) code and documentation was created by Artem Sokolov The methylated DNA stemness index (mDNAsi) code and documentation was created by Tathiane Maistro Malta This report was created by Michael Wells and Tiago Silva

# Install packages

To install the required packages to run the code below please execute the follwing code.
```{r eval = FALSE}
deps <- c("gelnet","dplyr","gdata","DT")
for(pkg in deps)  if (!pkg %in% installed.packages()) install.packages(pkg, dependencies = TRUE)
```

# Session info
```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(gelnet)
library(dplyr)
library(gdata)
library(DT)
```
```{r}
sessionInfo()
```
