---
title:  Trains a one-class dense model on the PCBC dataset
author: Artem Sokolov
output:
  html_document:
    highlight: tango
    theme: journal
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: no
editor_options:
  chunk_output_type: inline
bibliography: bibliography.bib
---
# Introduction

In the following workflow, we walk the reader through downloading the PCBC and TCGA PanCan33 datasets, training a stemness signature and applying it to score TCGA samples. By executing `main.train()` and `main.predict()` functions without any arguments will reproduce *mRNAsi* scores, as presented in the manuscript. The training function allows for further exploration of custom gene lists, provided through the `fnGenes` argument. The *EREG* scores in the manuscript were obtained using this functionality, by first identifying a set of differentially-methylated probes that distinguish stem cells and progenitors in PCBC, mapping those probes to the genes they regulate, and then limiting the training procedure to the resulting list of genes.

# Setup 

## R packages

Make sure you have the following packages downloaded and loaded into your library. 
For R synapse client please check this [article](http://docs.synapse.org/articles/getting_started.html).
We will retrieve data from [Synapse](https://www.synapse.org/), an open source data platform. 
A free account must be made before accessing synapse data, running the `synapseLogin` function will prompt you to login.

```{r,error = FALSE, eval = FALSE}
library(gelnet)
library(dplyr)
library(biomaRt)
library(synapseClient)
synapseLogin()
```

## Auxiliary functions

### Mapping ENSEMBL IDs to HUGO 

[Ensembl](www.ensembl.org) is a gene browser that will be used to gather information on the gene type and GeneID [@yates2015ensembl]. 
It can be accessed via R using the R/Bioconductor [biomaRt](http://bioconductor.org/packages/biomaRt/) package [@durinck2005biomart,@durinck2009mapping] 
which provides functions to connect to the Ensembl website and select the dataset you want with to work with. 
Using this package, we created the auxiliary function below to map human Ensemble gene IDs to HUGO Symbols.

# Methods


```{r, eval = FALSE}
# Maps ENSEMBL IDs to HUGO
# Use srcType = "ensembl_gene_id" for Ensembl IDs
# Use srcType = "entrezgene" for Entrez IDs
genes2hugo <- function( v, srcType = "ensembl_gene_id" )
{
    ## Retrieve the EMSEMBL -> HUGO mapping
    ensembl <- biomaRt::useMart( "ENSEMBL_MART_ENSEMBL", host="www.ensembl.org", dataset="hsapiens_gene_ensembl" )
    ID <- biomaRt::getBM( attributes=c(srcType, "hgnc_symbol"), filters=srcType, values=v, mart=ensembl )

    ## Make sure there was at least one mapping
    if( nrow(ID) < 1 ) top( "No IDs mapped successfully" )
    
    ## Drop empty duds
    j <- which( ID[,2] == "" )
    if( length(j) > 0 ) ID <- ID[-j,]
    stopifnot( all( ID[,1] %in% v ) )

    ID
}
```

## Train

In order to create our machine learning model, which will be used to make predictions on new data,
we will create a function called `main.train` (see complete code in the end of the section). 
Below we will explain each of its steps.

First, we will access synapse to retrieve the information. 
The function [synGet](https://www.rdocumentation.org/packages/synapseClient/versions/1.13-4/topics/synGet)
will be used to load and save RNA-seq data, while the function [synTableQuery](https://www.rdocumentation.org/packages/synapseClient/versions/1.13-4/topics/synTableQuery) is used to gather the meta data. 
In this case the download location is set to '/data/PCBC' but it can be changed to 
wherever you would like or it can be set to NULL and it will be stored in your cache.  

Data: `X` is a large matrix 13189 genes by 229 cell samples.  
```{r, eval = FALSE}
# Load RNAseq data
synRNA <- synGet( "syn2701943", downloadLocation = "/data/PCBC" )
```
```{r}
X <- read.delim( synRNA@filePath ) %>%
tibble::column_to_rownames( "tracking_id" ) %>% as.matrix
X[1:3,1:3]
```

Meta data: `Y` is a single variable data frame consisting of 301 observations.   

```{r}
# Retrieve metadata
synMeta <- synTableQuery( "SELECT UID, Diffname_short FROM syn3156503" )
Y <- synMeta@values %>%
  mutate( UID = gsub("-", ".", UID) ) %>%
  tibble::column_to_rownames( "UID" )
Y[1:4,]
```

Filter the labels from the meta data. Notice the change in format for y 
```{r}
# Retrieve the labels from the metadata
y <- Y[colnames(X),]

names(y) <- colnames(X)
# Fix the missing labels by hand
y["SC11.014BEB.133.5.6.11"] <- "EB"
y["SC12.039ECTO.420.436.92.16"] <- "ECTO"
  
## Drop the splice form ID from the gene names
v <- strsplit( rownames(X), "\\." ) %>% lapply( "[[", 1 ) %>% unlist()
  
rownames(X) <- v
head(y)
```

Call the `genes2hugo` function created before and use the labels from the RNA-seq data to map Ensembl gene ID and to HUGO symbols.
The result of this function is a data frame with two variables consisting of 12952 observations.

```{r}
# Map Ensembl IDs to HUGO
V <- genes2hugo( rownames(X) )
head(V)
```

Now change the row names of `X` from the gene name to the hgnc (HUGO Gene Nomenclature Committee) symbol. 
Notice that the dimensions of `X` change from 13189 rows to 12952 rows.
```{r}
X <- X[V[,1],]
rownames(X) <- V[,2]
X[1:3,1:3]
```

Reduce gene set to the provides list.
```{r, eval = FALSE}
if(!is.null(fnGenes)){
  vGenes <- read.delim( fnGenes, header=FALSE ) %>% as.matrix() %>% drop()
  VE <- genes2hugo( vGenes, "entrezgene" )
  X <- X[intersect( rownames(X), VE[,2] ),]
}
```

Find the mean center by subtracting the mean of each gene (`m`) from the RNA-seq data (`X`).
```{r}
m <- apply( X, 1, mean )
X <- X - m
X[1:3,1:3]
```

Identify stem cells and break up all samples into 2 groups:

* Stem cell(X.tr) -- 78 samples 
* not stem cell (X.bk) -- 151 samples 

```{r}
j <- which( y == "SC" )
X.tr <- X[,j]
X.tr[1:3,1:3]
```
```{r}
X.bk <- X[,-j]
X.bk[1:3,1:3]
```

Now we can begin to train the the one-class model with the [gelnet](https://www.rdocumentation.org/packages/gelnet/versions/1.2.1/topics/gelnet) function. 
The gelnet function can be used for Linear Regression, 
Binary Classification and One class Problems by using an iterative method called 
coordinated descent [@sokolov2016pathway].

```{r, eval=FALSE}
gelnet(X, y, l1, l2)
```

It has four main arguments described below:

* **X**: n by p matrix => transpose( X.r )
* **y**: `NULL` for one class models 
* **l1**: coefficient for the L1-norm penalty => 0 
* **l2**: coefficient for the L2-norm penalty => 1

Make sure you transpose the matrix so that the genes are listed as rows and samples as columns.
Then store the results as a tsv file (pcbc-stemsig.tsv). 
```{r, eval=F}
mm <- gelnet( t(X.tr), NULL, 0, 1 )
```
```{r, eval = FALSE}
write.table(mm$w, file = fnOut, sep = "\t", quote = FALSE, col.names = FALSE)
```


Leave-one-out cross validation will be used to test the accuracy of the model. 
Running this chunk will return a numeric vector containing the same size as the 
number of samples of the X.tr variable (in this case 78). If it is successful the returned values should all be equal to 1.

```{r, eval = F}
## Perform leave-one-out cross-validation
auc <- c()
for(i in 1:ncol(X.tr)){
  ## Train a model on non-left-out data
  X1 <- X.tr[,-i]
  m1 <- gelnet( t(X1), NULL, 0, 1 )
  
  ## Score the left-out sample against the background
  s.bk <- apply( X.bk, 2, function(z) {cor( m1$w, z, method="sp" )} )
  s1 <- cor( m1$w, X.tr[,i], method="sp" )
  
  ## AUC = P( left-out sample is scored above the background )
  auc[i] <- sum( s1 > s.bk ) / length(s.bk)
  cat( "Current AUC: ", auc[i], "\n" )
  cat( "Average AUC: ", mean(auc), "\n" )
}
```

### Entire code: `main.train` function

```{r, eval= FALSE}
## fnOut - filename of the output signature
## fnGenes - [optional] filename of the list of entrez ID to consider
main.train <- function( fnOut = "pcbc-stemsig.tsv", fnGenes = NULL )
{
  ## Load RNAseq data
  synRNA <- synGet( "syn2701943", downloadLocation = "/data/PCBC" )
  X <- read.delim( synRNA@filePath ) %>%
    tibble::column_to_rownames( "tracking_id" ) %>%
    as.matrix()
  
  ## Retrieve metadata
  synMeta <- synTableQuery( "SELECT UID, Diffname_short FROM syn3156503" )
  Y <- synMeta@values %>%
    mutate( UID = gsub("-", ".", UID) ) %>%
    tibble::column_to_rownames( "UID" )
  
  ## Retrieve the labels from the metadata
  y <- Y[colnames(X),]
  names(y) <- colnames(X)
  
  ## Fix the missing labels by hand
  y["SC11.014BEB.133.5.6.11"] <- "EB"
  y["SC12.039ECTO.420.436.92.16"] <- "ECTO"
  
  ## Drop the splice form ID from the gene names
  v <- strsplit( rownames(X), "\\." ) %>% lapply( "[[", 1 ) %>% unlist()
  rownames(X) <- v
  
  ## Map Ensembl IDs to HUGO
  V <- genes2hugo( rownames(X) )
  X <- X[V[,1],]
  rownames(X) <- V[,2]
  
  ## Reduce the gene set to the provided list (if applicable)
  if( is.null( fnGenes ) == FALSE )
  {
    vGenes <- read.delim( fnGenes, header=FALSE ) %>% as.matrix() %>% drop()
    VE <- genes2hugo( vGenes, "entrezgene" )
    X <- X[intersect( rownames(X), VE[,2] ),]
  }
  
  ## Mean-center the data
  m <- apply( X, 1, mean )
  X <- X - m
  
  ## Identify stem cell samples
  j <- which( y == "SC" )
  X.tr <- X[,j]
  X.bk <- X[,-j]
  
  ## Train a one-class model
  mm <- gelnet( t(X.tr), NULL, 0, 1 )
  
  ## Store the signature to a file
  write.table(mm$w, file = fnOut, sep = "\t", quote = FALSE, col.names = FALSE)
  
  ## Perform leave-one-out cross-validation
  auc <- c()
  for( i in 1:ncol(X.tr) )
  {
    ## Train a model on non-left-out data
    X1 <- X.tr[,-i]
    m1 <- gelnet( t(X1), NULL, 0, 1 )
    
    ## Score the left-out sample against the background
    s.bk <- apply( X.bk, 2, function(z) {cor( m1$w, z, method="sp" )} )
    s1 <- cor( m1$w, X.tr[,i], method="sp" )
    
    ## AUC = P( left-out sample is scored above the background )
    auc[i] <- sum( s1 > s.bk ) / length(s.bk)
    cat( "Current AUC: ", auc[i], "\n" )
    cat( "Average AUC: ", mean(auc), "\n" )
  }
  
  return(auc)
}
```


## Predict

In order to predict the classes to which unseen samples belong we will create a function called
`main.predict` (see complete code in the end of the section). Below we will explain each of its steps.

We start by using the [read.delim](http://rfunction.com/archives/1441) function 
to read the signature from the saved file (pcbc-stemsig.tsv) and store it as a variable `w`. 
`w` is a numeric vector the same size as the number of selected genes, in this case 12952.

```{r eval=FALSE}
fnSig = "pcbc-stemsig.tsv"
w <- read.delim(fnSig, header = FALSE, row.names = 1 ) %>% as.matrix() %>% drop()
```
```{r eval=FALSE}
w[1:10]
```

Again, we retrieve data from synapse. For this specific data you will need NIH approval to get access. 
Once you gather the data, create a data frame `X` and filter so that `X` contains genes 
that mapped and values from the signature gene set. 
`X` will be a large data frame with 11810 rows by 11070 columns.
```{r, eval = FALSE}
s <- synGet( "syn4976369", downloadLocation = "/data/pancan" )

# Auxiliary function: Reduces HUGO|POSITION gene IDs to just HUGO
f <- function( v ) unlist( lapply( strsplit( v, "\\|" ), "[[", 1 ) )

X <- read.delim( s@filePath, as.is=TRUE, check.names=FALSE ) %>%	## Read the raw values
     filter( !grepl( "\\?", gene_id ) ) %>%		## Drop genes with no mapping to HUGO
     mutate( gene_id = f( gene_id ) ) %>%		## Clip gene ids to HUGO
     filter( gene_id %in% names(w) )			## Reduce to the signature's gene set
X
```

If the data for SLC35E2 has multiple entries, we will filter it by and keep only the first one.

```{r, eval = FALSE}
j <- grep( "SLC35E2", X[,1] )
if( length(j) > 1 ) X <- X[-j[-1],]
```

Convert the data frame `X` to a matrix. Rows are Gene id's and columns are Gene Sets. 
Notice the the size of this matrix is now 11809 rows by 11069 columns. 
```{r, eval = FALSE}
rownames(X) <- NULL
X <- X %>% tibble::column_to_rownames( "gene_id" ) %>% as.matrix()
X[1:3,1:3]
```

Reduce the signature to the common set of genes.
```{r,eval = FALSE}
stopifnot( all( rownames(X) %in% names(w) ) )
w <- w[ rownames(X) ]
w[1:5]
```

Score the Matrix `X` using Spearman correlation. 
```{r, eval = FALSE}
s <- apply( X, 2, function(z) {cor( z, w, method="sp", use="complete.obs" )} )
s[1:5]
```

Scale the scores to be between 0 and 1
```{r, eval = FALSE}
s <- s - min(s)
s <- s / max(s)
s[1:5]
```

Then output scores to a file mRNA_StemScore.tsv.
```{r, eval=FALSE}
write.table( cbind(s), file=fnOut, sep="\t", quote=FALSE, col.names=FALSE )
```

### Entire code: `main.predict` function 
```{r, eval = FALSE}
## Uses the signature stored in fnSig to score PanCan33 data and stores the result to fnOut
main.predict <- function( fnSig = "pcbc-stemsig.tsv", fnOut = "mRNA_StemScore.tsv" )
{
  ## Load the signature
  w <- read.delim( fnSig, header=FALSE, row.names=1 ) %>% as.matrix() %>% drop()
  
  ## Reduces HUGO|POSITION gene IDs to just HUGO
  f <- function( v ) unlist( lapply( strsplit( v, "\\|" ), "[[", 1 ) )
  
  s <- synGet( "syn4976369", downloadLocation = "/data/pancan" )
  X <- read.delim( s@filePath, as.is=TRUE, check.names=FALSE ) %>%	## Read the raw values
    filter( !grepl( "\\?", gene_id ) ) %>%		## Drop genes with no mapping to HUGO
    mutate( gene_id = f( gene_id ) ) %>%		## Clip gene ids to HUGO
    filter( gene_id %in% names(w) )			## Reduce to the signature's gene set
  
  ## SLC35E2 has multiple entries with the same HUGO id
  ## Keep the first entry only
  j <- grep( "SLC35E2", X[,1] )
  if( length(j) > 1 )
    X <- X[-j[-1],]
  
  ## Convert to a matrix
  rownames(X) <- NULL
  X <- X %>% tibble::column_to_rownames( "gene_id" ) %>% as.matrix()
  
  ## Reduce the signature to the common set of genes
  stopifnot( all( rownames(X) %in% names(w) ) )
  w <- w[ rownames(X) ]
  
  ####### Score via Spearman correlation
  s <- apply( X, 2, function(z) {cor( z, w, method = "sp", use = "complete.obs" )} )
  
  ## Scale the scores to be between 0 and 1
  s <- s - min(s)
  s <- s / max(s)
  
  write.table(cbind(s), file = fnOut, sep = "\t", quote = FALSE, col.names = FALSE)
}
```



## Executing complete analysis

Once you have created all of the previous function (`main.train`, `main.predict`), 
create the `main` function which wraps the all them. 
This function will train and apply the full and reduced signatures. 
After you have have created all functions, run the `main` function to preform the full analysis. 

```{r, eval = FALSE}
main <- function()
{
  # Train a full signature, which will be saved to pcbc-stemsig.tsv
  main.train()
  
  # Apply the full signature to score the entire PanCan33 cohort
  main.predict()
}
```

# Conclusion 

We demonstrated how to derive a gene signature capable of detecting stem cell states and applied this signature to reproduce *mRNAsi*. The signature itself was stored into a file (`pcbc-stemsig.tsv` by default), allowing for additional downstream analyses, like the Gene Set Enrichment Analysis. The robustness of the signature was estimated through leave-one-out cross-validation that is automatically performed by the `main.train()` function. After stepping through the workflow, we encourage the reader to replace PanCan33 dataset with their own samples and modify `main.predict()` to derive the corresponding *mRNAsi*.

# References
